Hydraulic bore interaction with a column - A comparison between the solution of the shallow equation and experimental results

### 2015-03-12 doc/paper/paper.tex
add plots for waveheight after column and update paper.tex

The difference in shock shape can be explained by the difference between the shallow water equations that GeoClaw solves and the real physics of fluid. 
In shallow water equations, fluid viscosity is not taken into account, while it is significant in real physics.
This causes the shock remain its shape as it propagates in shallow water equation while smooth out in experiment.
Thus we see a jump in result from shallow water equation while a smooth increase in experiment result.
\par
This time interval is much smaller than that in Figure \ref{fig:waveheight_square_x=11.0} though data are not measured at exactly the same point for this two plots. 
This difference indicates that the column gives more resistance to shock waves in experiment than it does in shallow water equation model. 
\par
A simple analysis to this phenomenon is drawn as below.
The shallow water equations are essentially a system of hyperbolic PDEs. The solution depends on initial condition along characteristics in the spatio-temporal domain. 
Thus before the propagating shock arrives, all water downstream the shock has zero velocity. 
In other words, it even does not know there is a column before the shock hits the column.
This can be verified by checking the history of the wave height measured at exactly the same location, $0.04m$ ahead of the leading edge of the square column, but in different cases, one with a square column and one without.
This comparison is showed in Figure \ref{fig:waveheight_x11_comparison}. As we expected, there is no difference in arrival time of the shock wave.
In contrast, the Navier-Stokes equations, which better describes the real physics, are a system of PDEs of elliptic type. 
In real physics or in Navier-Stokes equations, when front of the shock wave gets close to the column, water downstream it and upstream column are affected (have nonzero velocity) and resist the comming of wave to some extent. 
This slows down the propagating speed of shock wave when it comes close to the column in real world and thus shifts the peak toward later time in the plots.
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.5\textwidth]{./plots/waveheight_square_x11}
    \caption{time history of wave height for case with square column}
    \label{fig:waveheight_square_x=11.0}
\end{figure}
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.5\textwidth]{./plots/waveheight_x11_comparison}
    \caption{comparison of time history of wave height for case with a square column and case without one}
    \label{fig:waveheight_x11_comparison}
\end{figure}
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.5\textwidth]{./plots/waveheight_square_x1102}
    \caption{time history of wave height for case with square column, at x = 11.02}
    \label{fig:waveheight_square_x=11.02}
\end{figure}
\par
Figure \ref{fig:waveheight_square_x=11.04} shows the time history of the wave height measured at the front face of the square column. 
The difference between the arrival time of the peak still exists as expected. After the peak arrives, the wave gradually becomes stable as indicated by a horizontal line in the plot. In this region the differences between GeoClaw and experiment are quite small. 
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.5\textwidth]{./plots/waveheight_square_x1104_largerTimeRange}
    \caption{time history of wave height for case with square column, at x = 11.04}
    \label{fig:waveheight_square_x=11.04}
\end{figure}

\section{Dam Break with cylinder column}\label{Sec:Cylinder}
Figure \ref{fig:waveheight_cylinder_x=11.0} shows the time history of the wave height measured at $0.04m$ ahead of the leading edge of cylinder column($0.1m$ ahead of the center of the column).
%Since the way GeoClaw treat a cell partly belong to cylinder and partly belong to tank bottom is to give that cell an average value of topography from cylinder and tank bottom based on area ratio, we get a quasi-circle actually instead of circle. Thus the result in this case does not agree well with the experiment result. 
%Future test can be conducted by setting a higher level for mesh refinement to get a better circle.
Similar feature as in the square column case Figure \ref{fig:waveheight_square_x=11.0} can be found.
However, the wave height at the second peak in this case is almost the same as that in square column case (around $0.22m$ in both case), while this feature is quite different for the experimental results between the two cases.
\par
Note that diameter of the cylinder is equal to half the length of the square column. Thus the square column can block the flow in in a more efficient way than cylinder does, since the cylinder has a smoother shape for the fluid to pass by. 
This explains why we see a lower peak of wave height as well as less steep curve for history of wave height in case with the cylinder in the experiment. 
In GeoClaw, a geometry is represented with quadrilateral cells. In the cylinder case, the cylinder is represented with square cells since our background mesh is simple structured quadrilateral mesh with square cells.
Based on this, we can deduce in GeoClaw, when the shock wave hits the cylinder, it is hitting onto many small faces vertical to flow direction at different locations. 
Thus it is similar to hitting onto one big face vertical to flow direction, which is the case with square column.  
This is why my get similar result in GeoClaw for case with square column and case with cylinder.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.5\textwidth]{./plots/waveheight_cylinder_x11}
    \caption{time history of wave height for case with cylinder column}
    \label{fig:waveheight_cylinder_x=11.0}
\end{figure}
\par
Figure \ref{fig:waveheight_cylinder_x=11.04} shows the time history of the wave height for the cylinder case. As we expect, the peak predicted by GeoClaw appears earlier but the peak values agree quite well with the experiment.
After the peak, GeoClaw gives a satisfied prediction of the wave height.
\begin{figure}[h!]
    \centering
    \includegraphics[width=0.5\textwidth]{./plots/waveheight_cylinder_x1104_largerTimeScale}
    \caption{time history of wave height for case with cylinder column, at x = 11.04}
    \label{fig:waveheight_cylinder_x=11.04}
\end{figure}

>>>>

The difference in shock shape can be explained by the difference between the shallow water equations that GeoClaw solves and the real physics of fluid. 
In shallow water equations, fluid viscosity is not taken into account, while it is significant in real physics.
This causes the shock remain its shape as it propagates in shallow water equation while smooth out in experiment.
Thus we see a jump in result from shallow water equation while a smooth increase in experiment result.
This time interval is much smaller than that in Figure \ref{fig:waveheight_square_x=11.0} though data are not measured at exactly the same point for this two plots. 
This difference indicates that the column gives more resistance to shock waves in experiment than it does in shallow water equation model. 
A simple analysis to this phenomenon is drawn as below.
The shallow water equations are essentially a system of hyperbolic PDEs. The solution depends on initial condition along characteristics in the spatio-temporal domain. 
Thus before the propagating shock arrives, all water downstream the shock has zero velocity. 
In other words, it even does not know there is a column before the shock hits the column.
This can be verified by checking the history of the wave height measured at exactly the same location, $0.04m$ ahead of the leading edge of the square column, but in different cases, one with a square column and one without.
This comparison is showed in Figure \ref{fig:waveheight_x11_comparison}. As we expected, there is no difference in arrival time of the shock wave.
In contrast, the Navier-Stokes equations, which better describes the real physics, are a system of PDEs of elliptic type. 
In real physics or in Navier-Stokes equations, when front of the shock wave gets close to the column, water downstream it and upstream column are affected (have nonzero velocity) and resist the comming of wave to some extent. 
This slows down the propagating speed of shock wave when it comes close to the column in real world and thus shifts the peak toward later time in the plots.
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=1\textwidth]{./plots/waveheight_square_x11}
        \caption{at x = 11.0 (0.04m ahead of front face of the square column)}
        \label{fig:waveheight_square_x=11.0}
    \end{subfigure}
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=1\textwidth]{./plots/waveheight_square_x1102}
        \caption{at x = 11.2 (0.02m ahead of front face of the square column)}
        \label{fig:waveheight_square_x=11.02}
    \end{subfigure}
    \caption{time history of wave height for case with a square column.}
%\begin{figure}[h!]
%    \centering
%    \includegraphics[width=0.5\textwidth]{./plots/waveheight_square_x11}
%    \caption{time history of wave height for case with square column}
%    \label{fig:waveheight_square_x=11.0}
%\end{figure}
Figure \ref{fig:waveheight_square_x=11.04} shows the time history of the wave height measured at the front face of the square column. 
The difference between the arrival time of the peak still exists as expected. After the peak arrives, the wave gradually becomes stable as indicated by a horizontal line in the plot. In this region the differences between GeoClaw and experiment are quite small. 
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=1\textwidth]{./plots/waveheight_x11_comparison}
        \caption{comparison between case with a square column and case without one}
        \label{fig:waveheight_x11_comparison}
    \end{subfigure}
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=1\textwidth]{./plots/waveheight_square_x1104_largerTimeRange}
        \caption{case with a square column with a larger time range at x = 11.04}
        \label{fig:waveheight_square_x=11.04}
    \end{subfigure}
    \caption{time history of wave height for square column case}
%\begin{figure}[h!]
%    \centering
%    \includegraphics[width=0.5\textwidth]{./plots/waveheight_x11_comparison}
%    \caption{comparison of time history of wave height for case with a square column and case without one}
%    \label{fig:waveheight_x11_comparison}
%\end{figure}
%\begin{figure}[h!]
%    \centering
%    \includegraphics[width=0.5\textwidth]{./plots/waveheight_square_x1102}
%    \caption{time history of wave height for case with square column, at x = 11.02}
%    \label{fig:waveheight_square_x=11.02}
%\end{figure}

%\begin{figure}[h!]
%    \centering
%    \includegraphics[width=0.5\textwidth]{./plots/waveheight_square_x1104_largerTimeRange}
%    \caption{time history of wave height for case with square column, at x = 11.04}
%    \label{fig:waveheight_square_x=11.04}
%\end{figure}

\section{Dam Break with cylinder column}\label{Sec:Cylinder}
Figure \ref{fig:waveheight_cylinder_x=11.0} shows the time history of the wave height measured at $0.04m$ ahead of the front point of cylinder column($0.1m$ ahead of the center of the column).
%Since the way GeoClaw treat a cell partly belong to cylinder and partly belong to tank bottom is to give that cell an average value of topography from cylinder and tank bottom based on area ratio, we get a quasi-circle actually instead of circle. Thus the result in this case does not agree well with the experiment result. 
%Future test can be conducted by setting a higher level for mesh refinement to get a better circle.
Similar feature as in the square column case Figure \ref{fig:waveheight_square_x=11.0} can be found.
However, the wave height at the second peak in this case is almost the same as that in square column case (around $0.22m$ in both case), while this feature is quite different for the experimental results between the two cases.
\par
Note that diameter of the cylinder is equal to half the length of the square column. 
Thus in experiment, the square column can block the flow in in a more efficient way than cylinder does, since the cylinder has a smoother shape for the fluid to pass by. 
This explains why we see a lower peak of wave height as well as less steep curve for history of wave height in case with the cylinder in the experiment. 
In GeoClaw, a geometry is represented with quadrilateral cells. 
In the cylinder case, the cylinder is represented with square cells since our background mesh is simple structured quadrilateral mesh with square cells.
Based on this, we can deduce in GeoClaw, when the shock wave hits the cylinder, it is hitting onto many small faces vertical to flow direction at different locations. 
In other words, there is no cell which has a normal vector of front face pointing toward oblique direction other than the direction fluid comes.  
Thus it is similar to hitting onto one big face vertical to flow direction, which is the case with square column.  
This is why we get similar result in GeoClaw for case with square column and case with cylinder.
\begin{figure}[h!]
    \centering
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=1\textwidth]{./plots/waveheight_cylinder_x11}
        \caption{at x = 11}
        \label{fig:waveheight_cylinder_x=11.0}
    \end{subfigure}
    \begin{subfigure}[b]{0.4\textwidth}
        \includegraphics[width=1\textwidth]{./plots/waveheight_cylinder_x1104_largerTimeScale}
        \caption{at x = 11.04}
        \label{fig:waveheight_cylinder_x=11.04}
    \end{subfigure}
    \caption{time history of wave height for cylinder column case}
\end{figure}

%\begin{figure}[h!]
%    \centering
%    \includegraphics[width=0.5\textwidth]{./plots/waveheight_cylinder_x11}
%    \caption{time history of wave height for case with cylinder column}
%    \label{fig:waveheight_cylinder_x=11.0}
%\end{figure}
\par
Figure \ref{fig:waveheight_cylinder_x=11.04} shows the time history of the wave height for the cylinder case. As we expect, the peak predicted by GeoClaw appears earlier but the peak values agree quite well with the experiment.
After the peak, GeoClaw gives a satisfied prediction of the wave height.
%\begin{figure}[h!]
%    \centering
%    \includegraphics[width=0.5\textwidth]{./plots/waveheight_cylinder_x1104_largerTimeScale}
%    \caption{time history of wave height for case with cylinder column, at x = 11.04}
%    \label{fig:waveheight_cylinder_x=11.04}
%\end{figure}
